---
title: "UFO sightings"
author: "collnell"
date: "November 30, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(knitr)
library(ggplot2)
library(dplyr)
se <- function(x) sd(x)/sqrt(length(x))
```

##dplyr: grammar of data manipulation    

```{r,eval=F}
install.packages(c("dplyr","RCurl"))
library(dplyr)
library(RCurl)
```  
###Main verbs  
- select() & filter()  
- group_by()  
- summarize() & mutate() 
- join()  

https://cran.r-project.org/web/packages/dplyr/dplyr.pdf


##UFO Reporting Center data  {.smaller}  
    
```{r}
nuforc<-read.csv("https://raw.githubusercontent.com/collnell/UFOs/master/ufos_2014_2016.csv")
str(nuforc)
```  
http://www.nuforc.org/  

##UFO sightings over time   {.smaller}
mutate() to create a new variable from existing variables  
summarize() to create a value from a vector (based on group_by levels)  
```{r}
ufos_date<-nuforc%>%
  mutate(Date = as.Date(paste(Year, Month, Day, sep="/"), format="%Y/%B/%d"))%>%
  group_by(Date, Month, Year)%>%
  na.omit()%>%
  summarize(UFOs = length(Day))
str(ufos_date)

```  
  
## UFO sightings  {.smaller}      
```{r}

ggplot(ufos_date, aes(x=Date, y=UFOs))+geom_point(size=1)+theme_minimal()+
  scale_x_date(date_labels= "%b-%Y")


```
```{r}
max(ufos_date$UFOs)
head(ufos_date[ufos_date$UFOs > 50, ], 10)


```

###Which US states have had the highest average monthly UFO sightings?  

## select: subset columns  {.smaller .build}  

```{r}
ufos <- select(nuforc, Year, Month, Day, Country, State, City, Shape, Summary)
head(ufos)
```  
or drop unwanted variables:  
```{r, eval=F}
ufos<-select(nuforc, -Duration, -Time)
```  

## filter: subset rows {.smaller .build} 

```{r}
ufos$Summary<-as.character(ufos$Summary)
aliens<-filter(ufos, grepl("alien", Summary))
head(aliens$Summary,15)

```  
## filter: subset rows {.smaller .build} 
Which US states have had the highest monthly UFO sightings?    
```{r}
unique(ufos$Country) #What countries are the sightings from?
length(unique(ufos$State)) #How many states have sightings?
```
Remove observations from Canada and NA states:    
```{r}
ufos_usa<-filter(ufos, Country == "USA", !is.na(ufos$State))

length(unique(ufos_usa$State)) 
```  

##pipelining with %>%    {.smaller .build}
- Execute multiple operations in given order   
- Increases readability with many commands 
- Ctrl + Shift + m  
  
Which US states have had the highest monthly UFO sightings?    
```{r}
ufos<-read.csv("https://raw.githubusercontent.com/collnell/UFOs/master/ufos_2014_2016.csv")%>%
  select(-Summary, -Duration, -Time)%>%
  filter(Country == "USA", !is.na(State))
```   


##summarize()  {.smaller .build} 
Takes a vector of values and return a single value for designated groups in group_by()  

```{r}
ufos_state<-ufos%>%
  group_by(State, Year, Month)%>%
  summarize(UFO_n = length(Day))
head(ufos_state)
``` 


##Exercise:  {.build}
calculate & plot monthly UFO sightings by state  
```{r}
se <- function(x) sd(x)/sqrt(length(x)) #hint
```  


```{r}
ufos_state <- ufos%>%
  group_by(State, Year, Month)%>%
  summarize(UFO_n = length(Day))%>% #count number of observations
  group_by(State)%>%
  summarize(UFO_mean = mean(UFO_n), UFO_se = se(UFO_n))%>% 
  arrange(UFO_mean)# sort rows in descending order
```    
arrange(): sort rows by variables  

##Monthly UFO sightings   {.smaller }   

```{r,echo=F, fig.width=8, fig.height=4}
top_states<-top_n(ufos_state, n=25)
ggplot(top_states, aes(x = reorder(State, UFO_mean), y = UFO_mean))+
  geom_point(size = 3)+
  geom_errorbar(aes(ymin = UFO_mean-UFO_se, ymax=UFO_mean+UFO_se), width= .2)+
  theme_minimal()+labs(x="State", y="Montly UFO sightings")
```  
  
```{r,eval=F}
ggplot(top_states, aes(x = reorder(State, UFO_mean), y = UFO_mean))+geom_point(size=3)+
  geom_errorbar(aes(ymin = UFO_mean-UFO_se, ymax=UFO_mean+UFO_se), width= .2)+
  theme_minimal()+labs(x="State", y="Montly UFO sightings")
```  

##join data   {.smaller}  
Merge UFO data with state census data to account for population size in UFO sightings  

```{r,eval=T,message=F, warning=F}
months<-as.ordered(c("January","February","April","May","June","July",
                     "August","September","October","November","December"))  
state_census<-read.csv("https://raw.githubusercontent.com/collnell/UFOs/master/states.csv") #census data

ufos<-ufos%>%
  filter(Country == "USA", !is.na(State), Month %in% months)%>%
  left_join(state_census, by=c("State"="state"))%>%
  group_by(Month, region, division, State, population, area)%>%
  summarize(UFOs = length(Shape))
head(ufos)  
```  
## mutate()  
Make new variables from existing variables  
```{r,eval=F}
ufos.month<-ufos%>%
  mutate(UFO_pop = UFOs/population*1000000) ##UFOs per million ppl
```  
mutate_all(funs())  
- same structure but takes a list of functions and applies them to all numeric variables  

summarize_all(funs(mean, se, length))  
- calculates mean, se, and n for all numeric variables


##Exercise {.smaller .build}  
Visualize UFO sightings using ggplot2   
```{r}
ufos.month<-ufos%>%
  mutate(UFO_pop = UFOs/population*1000000)%>% 
  group_by(Month,division)%>%
  summarize(UFO_month = mean(UFO_pop)) #sightings per month

```  
  
```{r, eval=F}
ggplot(ufos.month, aes(x=Month, y=UFO_month,group=division, color=division))+
  geom_point(size=2)+
  theme(axis.text.x=element_text(angle=90))+
  theme_minimal()+
  geom_line(size=1)+
  scale_x_discrete(name="Month", labels=months)+
  labs(x="", y="UFO sightings a month (for every 1,000,000 people)")

```  
  
##Montly UFO sightings  
  
```{r, eval=T, echo=F}
ggplot(ufos.month, aes(x=Month, y=UFO_month,group=division, color=division))+
  geom_point(size=2)+
  theme(axis.text.x=element_text(angle=90))+
  theme_minimal()+
  geom_line(size=1)+
  scale_x_discrete(name="Month", labels=months)+
  labs(x="", y="UFO sightings a month (for every 1,000,000 people)")

``` 

##casting data to wide form {.smaller .build}   
dcast() takes variables and creates new columns based on levels  
Cast dataframe by UFO shape  
```{r,warning=F}
library(reshape2)

ufos<-read.csv("ufos_2014_2016.csv")%>%
  left_join(state_census, by=c("State"="state"))%>%
  group_by(Shape, region)%>%
  na.omit()%>%
  summarize(UFOs = length(Day))%>%
  dcast(region~Shape, value.var="UFOs")
colnames(ufos)

```  
##melting data to long form  

In 'reshape2' package  
melt()  
```{r}
#reverse
ufo_long<-melt(ufos, id.vars="region", variable.name="Shape")
str(ufo_long)

```


##dplyr cheat sheet  
  
https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf   
    
